Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-07 writeup/conStruct.tex
minor edits for clarity

We estimate the values of the parameters of the model using a Bayesian approach.
Acknowledging the dependence of the parametric covariance matrix $\Omega$ on its constituent parameters
$w,\alpha,\phi,\eta,\gamma$ and on the (observed) geographic distances $D$ 
with the notation $\Omega(w,\alpha,\phi,\eta,\gamma,D)$,
we denote the posterior probability density of the parameters as:
\begin{equation}
P\left( w,\alpha,\phi,\eta,\gamma \;	| \; \widehat{\Omega}, L \right) \propto
P\left(\widehat{\Omega} \; | \; \Omega(w,\alpha,\phi,\eta,\gamma,D) \right)
P(w)P(\alpha)P(\phi)P(\eta)P(\gamma) ,
\end{equation}
where $P(w)$, $P(\alpha)$, $P(\phi)$, $P(\eta)$, and $P(\gamma)$, are prior distributions.
All parameters are given (half-)Gaussian priors 
except for $\alpha_2$, which is uniform on $(0,2)$, 
and $w$, for which we use an independent Dirichlet of dimension $K$ for each individual
(see Table \ref{tab:param_prior_tab} for specifics).
Parameters are independent between layers.
We use Hamiltonian Monte Carlo as implemented in STAN \citep{stan, NUTS, stan_lib, rstan} 
to estimate the posterior distribution on the parameters.
We also provide an R package \citep{R} called \texttt{conStruct} that functions as a wrapper 
around this inference machinery.

@@

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=2in]{figs/sims/sim_xvals.pdf}}
	\caption{
	Cross-validation results for data simulated under $K=1$, $K=2$, and $K=3$, 
	comparing the spatial and nonspatial \texttt{conStruct} models run with $K=1$ through 7.  
	The inset plots zoom in on cross-validation results outlined in the dotted boxes.
    }\label{sim_xvals}
\end{figure}

@@

Our data, originally published by \cite{geraldes_etal_2014}, 
consist of 434 individuals sampled from 35 drainages 
and genotyped at just over 33,000 loci (map of the sampling shown in Fig \ref{populus_map}).
The data were generated using RAD-seq, 
and showed a strong pattern of bias in allelic dropout 
(the majority of missing data were from drainages with only \textit{Populus balsamifera} individuals).
To ameliorate some of the problems that arise when there is a strong bias in which data are missing, 
we dropped loci for which any data were missing, 
resulting in just over 20,200 loci retained for analysis.  
We then analyzed these data, grouped by drainage, using both the spatial and nonspatial \texttt{conStruct} models, 
using $K = 1$ through 7,
and compared these models using cross-validation.
The results of these analyses are shown in 
Figs \ref{populus_xvals} and \ref{populus_pies}, 
as well as in Figs \ref{populus_sp_pies} - \ref{populus_laycon} in the Supplementary Materials.

@@

All models with $K>1$ assigned the majority of each species
to distinct layers, with some populations drawing ancestry from both layers.
This provides strong support for discrete population structure between the two species,
with some admixture,
rather than a single, continuous cline of ancestry.
At all values of $K>1$, discrete population structure was mostly partitioned along species lines; 
at values of $K$ above 2, further discrete substructure was inferred within the \textit{P. trichocarpa} samples, 
while \bals{} samples displayed no substructure up to $K=7$.
This lack of substructure within the sampled \bals{} populations is in agreement with results from 
\citet{keller_etal_2010}, who found that \bals{} in the region is characterized by relatively low diversity, 
and probably represents a recent, postglacial expansion.

@@

At all values of $K$, the spatial model was preferred over the nonspatial model (Fig \ref{populus_xvals}).
Most of this signal seems to derive from the \textit{P. trichocarpa} samples:
covariance within predominantly \tri{} layers decays more quickly with distance
than within the majority \bals{} layer, as seen in Fig \ref{populus_sp_clst_covs}.  
This indicates that, although isolation by distance is a feature of the dataset as a whole, 
it's magnitude varies considerably across biological units - 
an aspect of the data \texttt{conStruct} is able capture well.

@@

Our data, originally published by \cite{Puckett2015}, consist of 95 individuals 
sampled across the United States and on the West coast of Canada,
genotyped at just under 22,000 bi-allelic loci.
The distribution of missing data across these individuals was uneven, 
with a few individuals representing most of the missing data, 
so we removed individuals with greater than 4\% missing data, 
resulting in a final dataset of 78 individuals.
We then analyzed these data, treating individuals as the unit of analysis, 
using both the spatial and nonspatial \texttt{conStruct} models with a $K$ of between 1 and 7, 
and compared these models using cross-validation.
The results of these analyses are shown in Figs \ref{bear_xvals} - \ref{bear_laycon}, 
as well as in Figs \ref{bear_sp_pies} - \ref{bear_nsp_clst_covs} in the Supplementary Materials.

@@

Across all values of $K$ for which we ran \texttt{conStruct},
we see strong support for the spatial model over the nonspatial model (Fig \ref{bear_xvals}),
indicating the isolation by distance seems to be a prominent feature of the data.
This pattern may resolve a discrepancy between our results and 
previous analyses that split Alaskan and British Columbian bears out 
into their own cluster with an inferred Beringian glacial refugium
\citep{Byun1997,Stone2000,Puckett2015}.
Our model, which explicitly incorporates a spatial decay of relatedness, 
allows individuals sampled at a large remove to belong to the same layer
instead of splitting these individuals out into successive clusters.

@@

\begin{figure}
	\centering
		\subcaptionbox{$K=2$\label{bear_sp_pies:K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp2.pdf}}
		\subcaptionbox{$K=3$\label{bear_sp_pies:K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp3.pdf}}
		\subcaptionbox{$K=4$\label{bear_sp_pies:K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp4.pdf}}
		\subcaptionbox{$K=5$\label{bear_sp_pies:K5}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp5.pdf}}
		\subcaptionbox{$K=6$\label{bear_sp_pies:K6}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp6.pdf}}
		\subcaptionbox{$K=7$\label{bear_sp_pies:K7}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp7.pdf}}
	\caption{
	Map of admixture proportions estimated for the \textit{bear} dataset 
	using the spatial model for $K=2$ through 7.
    }\label{bear_sp_pies}
\end{figure}

\begin{figure}
	\centering
		\subcaptionbox{$K=2$\label{bear_nsp_pies:K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp2.pdf}}
		\subcaptionbox{$K=3$\label{bear_nsp_pies:K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp3.pdf}}
		\subcaptionbox{$K=4$\label{bear_nsp_pies:K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp4.pdf}}
		\subcaptionbox{$K=5$\label{bear_nsp_pies:K5}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp5.pdf}}
		\subcaptionbox{$K=6$\label{bear_nsp_pies:K6}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp6.pdf}}
		\subcaptionbox{$K=7$\label{bear_nsp_pies:K7}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp7.pdf}}
	\caption{
	Map of admixture proportions estimated for the \textit{bear} dataset 
	using the nonspatial model for $K=2$ through 7.
    }\label{bear_nsp_pies}
\end{figure}

>>>>

We estimate the values of the parameters of the model using a Bayesian approach.
Acknowledging the dependence of the parametric covariance matrix $\Omega$ on its constituent parameters
$w,\alpha,\phi,\eta,\gamma$ and on the (observed) geographic distances $D$ 
with the notation $\Omega(w,\alpha,\phi,\eta,\gamma,D)$,
we denote the posterior probability density of the parameters as:
\begin{equation}
P\left( w,\alpha,\phi,\eta,\gamma \;	| \; \widehat{\Omega}, L \right) \propto
P\left(\widehat{\Omega} \; | \; \Omega(w,\alpha,\phi,\eta,\gamma,D) \right)
P(w)P(\alpha)P(\phi)P(\eta)P(\gamma) ,
\end{equation}
where $P(w)$, $P(\alpha)$, $P(\phi)$, $P(\eta)$, and $P(\gamma)$, are prior distributions.
All parameters are given (half-)Gaussian priors 
except for $\alpha_2$, which is uniform on $(0,2)$, 
and $w$, for which we use an independent Dirichlet of dimension $K$ for each individual
(see Table \ref{tab:param_prior_tab} for specifics).
Parameters are independent between layers.
We use Hamiltonian Monte Carlo as implemented in STAN \citep{stan, NUTS, stan_lib, rstan} 
to estimate the posterior distribution on the parameters.
We also provide an R package \citep{R} called \texttt{conStruct} 
(for ``continuous structure") that functions as a wrapper around this inference machinery.

@@

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=2in]{figs/sims/sim_xvals.pdf}}
	\caption{
	Cross-validation results for data simulated under $K=1$, $K=2$, and $K=3$, 
	comparing the spatial and nonspatial \texttt{conStruct} models 
	(in blue and green, respectively) run with $K=1$ through 7.  
	The inset plots zoom in on cross-validation results outlined in the dotted boxes.
    }\label{sim_xvals}
\end{figure}

@@

Our data, originally published by \cite{geraldes_etal_2014}, 
consist of 434 individuals sampled from 35 drainages 
and genotyped at just over 33,000 loci (map of the sampling shown in Fig \ref{populus_map}).
The data were generated using RAD-seq, 
and showed a strong pattern of bias in allelic dropout 
(the majority of missing data were from drainages with only \textit{Populus balsamifera} individuals).
To ameliorate some of the problems that arise when there is a strong bias in which data are missing, 
we dropped loci for which any data were missing, 
resulting in just over 20,200 loci retained for analysis.  
We then analyzed these data, grouped by drainage, using both the spatial and nonspatial \texttt{conStruct} models, 
using $K = 1$ through 7,
and compared these models using cross-validation.
The results of these analyses are shown in 
Figs \ref{populus_pies} and \ref{populus_xvals}, 
as well as in Figs \ref{populus_sp_pies} - \ref{populus_laycon} in the Supplementary Materials.

@@

All models with $K>1$ assigned the majority of each species
to distinct layers, with some populations drawing ancestry from multiple layers.
This provides strong support for discrete population structure between the two species,
with some admixture,
rather than a single, continuous cline of ancestry.
At all values of $K>1$, discrete population structure was mostly partitioned along species lines; 
at values of $K$ above 2, further discrete substructure was inferred within the \textit{P. trichocarpa} samples, 
while \bals{} samples displayed no substructure up to $K=7$.
This lack of substructure within the sampled \bals{} populations is largely in agreement with results from 
\citet{keller_etal_2010}, who found that \bals{} in the region is characterized by relatively low diversity, 
and probably represents a recent, postglacial expansion.

@@

At all values of $K$, the spatial model was preferred over the nonspatial model (Fig \ref{populus_xvals}).
Most of this signal seems to derive from the \textit{P. trichocarpa} samples:
covariance within predominantly \tri{} layers decays more quickly with distance
than within the majority \bals{} layer, as seen in Fig \ref{populus_sp_clst_covs}.  
This indicates that, although isolation by distance is a feature of the dataset as a whole, 
its magnitude varies considerably across biological units - 
an aspect of the data \texttt{conStruct} is able capture well.

@@

Our data, originally published by \cite{Puckett2015}, consist of 95 individuals 
sampled across the United States and on the West coast of Canada,
genotyped at just under 22,000 bi-allelic loci.
The distribution of missing data across these individuals was uneven, 
with a few individuals representing most of the missing data, 
so we removed individuals with greater than 4\% missing data, 
resulting in a final dataset of 78 individuals.
We then analyzed these data, treating individuals as the unit of analysis, 
using both the spatial and nonspatial \texttt{conStruct} models with a $K$ of between 1 and 7, 
and compared these models using cross-validation.
The results of these analyses are shown in Figs \ref{bear_K3} - \ref{bear_laycon}, 
as well as in Figs \ref{bear_sp_pies} - \ref{bear_nsp_clst_covs} in the Supplementary Materials.

@@

Across all values of $K$ for which we ran \texttt{conStruct},
we see strong support for the spatial model over the nonspatial model (Fig \ref{bear_xvals}),
indicating the isolation by distance seems to be a prominent feature of the data.
This pattern may resolve a discrepancy between our results and 
previous analyses that split Alaskan and British Columbian bears out 
into their own cluster with an inferred Beringian glacial refugium
\citep{Byun1997,Stone2000,Puckett2015}.
Our model, which explicitly incorporates a spatial decay of relatedness, 
allows somewhat genetically differentiated individuals 
that are sampled far from one another to belong to the same layer,
instead of splitting these individuals out into successive clusters 
(e.g., Fig \ref{bear_sp_pies:K5} vs \ref{bear_nsp_pies:K5}).

@@

\begin{figure}
	\centering
		\subcaptionbox{$K=2$\label{bear_sp_pies:K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp2.pdf}}
		\subcaptionbox{$K=3$\label{bear_sp_pies:K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp3.pdf}}
		\subcaptionbox{$K=4$\label{bear_sp_pies:K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp4.pdf}}
		\subcaptionbox{$K=5$\label{bear_sp_pies:K5}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp5.pdf}}
		\subcaptionbox{$K=6$\label{bear_sp_pies:K6}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp6.pdf}}
		\subcaptionbox{$K=7$\label{bear_sp_pies:K7}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_sp7.pdf}}
	\caption{
	Map of admixture proportions estimated for the bear dataset 
	using the spatial model for $K=2$ through 7.
    }\label{bear_sp_pies}
\end{figure}

\begin{figure}
	\centering
		\subcaptionbox{$K=2$\label{bear_nsp_pies:K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp2.pdf}}
		\subcaptionbox{$K=3$\label{bear_nsp_pies:K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp3.pdf}}
		\subcaptionbox{$K=4$\label{bear_nsp_pies:K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp4.pdf}}
		\subcaptionbox{$K=5$\label{bear_nsp_pies:K5}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp5.pdf}}
		\subcaptionbox{$K=6$\label{bear_nsp_pies:K6}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp6.pdf}}
		\subcaptionbox{$K=7$\label{bear_nsp_pies:K7}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/bears/bear_nsp7.pdf}}
	\caption{
	Map of admixture proportions estimated for the bear dataset 
	using the nonspatial model for $K=2$ through 7.
    }\label{bear_nsp_pies}
\end{figure}
